---
title: "ML1_Final_EDA_ShinyApp_included"
date: "2024-05-26"
output: 
  html_document:
    df_print: paged
    theme: "sandstone"
    css: "css/style.css"
runtime: shiny
---

# 1. Libraries

This section loads the necessary R packages for data manipulation,
visualization, and other tasks. If a package is not installed, it
installs it first and then loads it.

```{r setup, include=FALSE}
# Load required libraries
# dplyr: For data manipulation
if (!require(dplyr, quietly = TRUE)) {
  install.packages("dplyr")
  library(dplyr)
}

# stringr: For string manipulation
if (!require(stringr, quietly = TRUE)) {
  install.packages("stringr")
  library(stringr)
}

# plotly: For interactive plots
if (!require(plotly, quietly = TRUE)) {
  install.packages("plotly")
  library(plotly)
}

# ggplot2: For data visualization
if (!require(ggplot2, quietly = TRUE)) {
  install.packages("ggplot2")
  library(ggplot2)
}

# readxl: For reading Excel files
if (!require(readxl, quietly = TRUE)) {
  install.packages("readxl")
  library(readxl)
}

# readr: For reading data
if (!require(readr, quietly = TRUE)) {
  install.packages("readr")
}
library(readr)

# treemapify: For creating treemaps
if (!require(treemapify, quietly = TRUE)) {
  install.packages("treemapify")
}
library(treemapify)

# scales: For scale functions
if (!require(scales, quietly = TRUE)) {
  install.packages("scales")
}
library(scales)

# patchwork: For combining ggplots
if (!require(patchwork, quietly = TRUE)) {
  install.packages("patchwork")
}
library(patchwork)

# gridExtra: For arranging plots
if (!require(gridExtra, quietly = TRUE)) {
  install.packages("gridExtra")
}
library(gridExtra)

# rcolorbrewer
if (!require(RColorBrewer, quietly = TRUE)) {
  install.packages("RColorBrewer")
}
library(RColorBrewer)

 # forcats
if (!require(forcats, quietly = TRUE)) {
  install.packages("forcats")
}
library(forcats)

 # shiny
if (!require(shiny, quietly = TRUE)) {
  install.packages("shiny")
}
library(shiny)

 # GAM
if (!require(mgcv, quietly = TRUE)) {install.packages("mgcv")}
library(mgcv)

if (!require(effects, quietly = TRUE)) {install.packages("effects")}
library(effects)
```

# 2. Dataset

This section for loading the dataset from the provided URL and
displaying a summary of the data.

```{r loading, include=FALSE}
# Load dataset
df <- read.csv("kul100od1001.csv")
```

Here are some of the key attributes:

StichtagDatJahr: Year of the data record HalterId: Identifier for the
pet owner AlterV10Cd, AlterV10Lang, AlterV10Sort: Codes, descriptions,
and sorting for the age group of the pet owner SexCd, SexLang, SexSort:
Codes, descriptions, and sorting for the gender of the pet owner
KreisCd, KreisLang, KreisSort: Codes, descriptions, and sorting for the
district QuarCd, QuarLang, QuarSort: Codes, descriptions, and sorting
for the quarter Rasse1Text, Rasse2Text: Primary and secondary breed
texts of the pet RasseMischlingCd, RasseMischlingLang,
RasseMischlingSort: Codes, descriptions, and sorting for whether the pet
is a mixed breed RassentypCd, RassentypLang, RassentypSort: Codes,
descriptions, and sorting for the type of breed GebDatHundJahr: Year of
birth of the pet AlterVHundCd, AlterVHundLang, AlterVHundSort: Codes,
descriptions, and sorting for the age group of the pet SexHundCd,
SexHundLang, SexHundSort: Codes, descriptions, and sorting for the
gender of the pet HundefarbeText: Color of the pet AnzHunde: Number of
dogs

# 3. Translation

In this section, we duplicate and rename the dataframe df as df_EN for
the English version. Then, translations for column names in English are
defined. Following this, a function is employed to replace multiple
patterns at once for content translation. Patterns and replacements for
content translation, including translations for age groups, sexes, breed
types, and dog colors, are defined. After applying the translation
function across all columns, dog colors are also translated.

```{r translation, include=FALSE}

# Duplicate and rename df for English version
df_EN <- df

# Define translations for column names in English
colnames(df_EN) <- c("KeyDateYear", "DataStatusCd", "OwnerId", "OwnerAgeGroupCd", "OwnerAgeGroup", "OwnerAgeGroupSort", "OwnerSexCd", "OwnerSex", "OwnerSexSort", "DistrictCd", "District", "DistrictSort", "QuarCd", "Quar", "QuarSort", "PrimaryBreed", "SecondaryBreed", "MixedBreedCd", "MixedBreed", "MixedBreedSort", "BreedTypeCd", "BreedType", "BreedTypeSort", "DogBirthYear", "DogAgeGroupCd", "DogAgeGroup", "DogAgeGroupSort", "DogSexCd", "DogSex", "DogSexSort", "DogColor", "NumberOfDogs")

# Define a function to replace multiple patterns at once
replace_patterns <- function(text, patterns, replacements) {
  for (i in seq_along(patterns)) {
    text <- str_replace_all(text, patterns[i], replacements[i])
  }
  return(text)
}

# Define patterns and replacements for content translation
patterns <- c("- bis ", "-Jährige", "männlich", "weiblich", "Keine", "Unbekannt", "Rassehund", "Mischling, beide Rassen bekannt", "Mischling, sekundäre Rasse unbekannt", "Mischling, beide Rassen unbekannt", "Kleinwüchsig", "Rassentypenliste I", "Rassentypenliste II")
replacements <- c(" to ", " years old", "male", "female", "none", "Unknown", "Pedigree dog", "Mixed breed, both breeds known", "Mixed breed, secondary breed unknown", "Mixed breed, both breeds unknown", "Small stature", "Breed type list I", "Breed type list II")

# Apply the function across all columns
df_EN[] <- lapply(df_EN, function(x) replace_patterns(x, patterns, replacements))

# Color translation - can be further customized based on your dataset
color_patterns <- c("schwarz", "braun", "weiss", "grau", "silber", "rot", "gelb", "hell", "dunkel", "gestromt", "schimmel", "zweifarbig", "dreifarbig", "vierfarbig", "gemischt", "meliert", "hirschrot mit Maske", "löwenfarbig")
color_replacements <- c("black", "brown", "white", "gray", "silver", "red", "yellow", "light", "dark", "brindle", "mold", "2 colors", "3 colors", "4 colors", "mixed", "mottled", "stag red with mask", "lion-colored")

# Translate dog colors
df_EN$DogColor <- replace_patterns(df_EN$DogColor, color_patterns, color_replacements)

# Optional: Print unique values to check the translations
print(unique(df_EN$MixedBreed))
print(unique(df_EN$BreedType))
print(unique(df_EN$DogColor))

head(df_EN)

write.csv(df_EN, "df_EN.csv", row.names = FALSE)
rm(df)
```

# 4. Unique Owner IDs

In the following R code snippet, we implement a method to distinguish
unique OwnerId values within our dataset. By marking the initial
occurrence of each OwnerId as unique, we facilitate further analyses
that may require the identification of distinct entries.

```{r df_unique_OwnerId, include=FALSE}
df_EN$unique_OwnerId <- !duplicated(df_EN$OwnerId)
dim(df_EN)

```

# 5. Refined Dataset

The R code below demonstrates the process of extracting a subset of
relevant columns from our comprehensive dataset df_EN, thereby creating
a streamlined DataFrame, new_df. This subset includes essential fields
such as KeyDateYear, OwnerId, and details regarding the dogs such as
PrimaryBreed and DogBirthYear. Additionally, the code converts the
NumberOfDogs column from its original format to a numeric type, ensuring
that subsequent data analysis can utilize numerical operations.

```{r new_df, include=FALSE}
# Create a new DataFrame with selected columns and convert 'NumberOfDogs' to numeric
new_df <- df_EN %>%
  select(KeyDateYear, OwnerId, OwnerAgeGroup, OwnerSex, DistrictSort, QuarCd, PrimaryBreed, SecondaryBreed, MixedBreed, BreedType, DogBirthYear,    DogSex, NumberOfDogs, unique_OwnerId) %>%
  mutate(NumberOfDogs = as.numeric(as.character(NumberOfDogs)))

dim(new_df)
str(new_df)

```

# 6. ML Models


....


## 6.6. How do dog breed popularity trends evolve over time?
(GAM use: Model the non-linear trend of various dog breeds' popularity
over the years to predict future trends. This can help PetParadise to
anticipate demand for breed-specific products.)

### 6.6.1. All ditrict together

```{r GAM for 5 Top breed  - ALL DISTRICT TOGETHER, echo=FALSE, message=FALSE, warning=FALSE}

# Ensure correct type
new_df$KeyDateYear <- as.numeric(as.character(new_df$KeyDateYear))
new_df$PrimaryBreed <- factor(new_df$PrimaryBreed)


# Aggregating data to get count of dogs per breed per year (ignoring district)
breed_trends <- new_df %>%
  group_by(KeyDateYear, PrimaryBreed) %>%
  summarise(BreedCount = sum(NumberOfDogs), .groups = 'drop') %>%
  filter(PrimaryBreed != "Unknown" & PrimaryBreed %in% c("Chihuahua", "Labrador Retriever", "Yorkshire Terrier", "Jack Russel Terrier", "Malteser")) %>%
  as.data.frame()

# Check the structure of the aggregated data
str(breed_trends)

# Ensure no NA values in the key variables
breed_trends <- breed_trends %>%
  filter(!is.na(BreedCount), !is.na(KeyDateYear))

#R-sq.(adj) =  0.994   Deviance explained = 99.6%
gam_model <- gam(BreedCount ~ te(KeyDateYear, by = PrimaryBreed, k = 9) + s(PrimaryBreed, bs = "re"),
                 data = breed_trends, method = "REML", family = poisson(link = "log"))


# Prepare new data for prediction
new_data <- data.frame(
  KeyDateYear = rep(seq(2015, 2030, by = 1), times = 5),
  PrimaryBreed = factor(rep(c("Chihuahua", "Labrador Retriever", "Yorkshire Terrier", "Jack Russel Terrier", "Malteser"), each = 16),
                        levels = c("Chihuahua", "Labrador Retriever", "Yorkshire Terrier", "Jack Russel Terrier", "Malteser"))
)

# Predict counts for the new dataset
new_data$predicted_breed_count <- predict(gam_model, newdata = new_data, type = "response")

# Function to assign colors to breeds
assign_colors <- function(data) {
  breeds <- unique(data$PrimaryBreed)
  colors <- scales::hue_pal()(length(breeds))
  setNames(colors, breeds)
}
breed_color_map <- assign_colors(new_data)

# Plot the results with real data points
ggplot() +
  geom_line(data = new_data, aes(x = KeyDateYear, y = predicted_breed_count, color = PrimaryBreed), size = 1) +
  geom_point(data = breed_trends, aes(x = KeyDateYear, y = BreedCount, color = PrimaryBreed), size = 2) +
  #ylim(0, 600) +
  labs(x = "Year", y = "Count", title = "Yearly Trend of Predicted and Actual Breed Counts (Summed across all Districts)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top") +
  guides(color = guide_legend(title = "Breed", title.position = "top"))

# Summary and diagnostic plots of the combined model
summary(gam_model)
#plot.gam(gam_model)

```

### 6.6.2. Per district

```{r GAM for 5 Top breed  - Per DISTRICT, echo=FALSE, message=FALSE, warning=FALSE}

# Ensure correct type
new_df$KeyDateYear <- as.numeric(as.character(new_df$KeyDateYear))
new_df$PrimaryBreed <- factor(new_df$PrimaryBreed)
new_df$DistrictSort <- factor(new_df$DistrictSort, levels = 1:12)

# Aggregating data to get count of dogs per breed per year per district
breed_trends <- new_df %>%
  group_by(KeyDateYear, PrimaryBreed, DistrictSort) %>%
  summarise(BreedCount = sum(NumberOfDogs), .groups = 'drop') %>%
  filter(PrimaryBreed != "Unknown" & 
         PrimaryBreed %in% c("Chihuahua", "Labrador Retriever", "Yorkshire Terrier", "Jack Russel Terrier", "Malteser")) %>%
  as.data.frame()

# Check the structure of the aggregated data
str(breed_trends)

# Ensure no NA values in the key variables
breed_trends <- breed_trends %>%
  filter(!is.na(BreedCount), !is.na(KeyDateYear), !is.na(DistrictSort))

# Prepare new data for prediction for each district
new_data <- expand.grid(
  KeyDateYear = seq(2015, 2030, by = 1),
  PrimaryBreed = factor(c("Chihuahua", "Labrador Retriever", "Yorkshire Terrier", "Jack Russel Terrier", "Malteser"),
                        levels = c("Chihuahua", "Labrador Retriever", "Yorkshire Terrier", "Jack Russel Terrier", "Malteser")),
  DistrictSort = factor(1:12, levels = 1:12)
)

# Function to fit the GAM model and make predictions for a specific district
fit_gam_and_predict <- function(district) {
  district_data <- breed_trends %>% filter(DistrictSort == district)
  
  # Check if there are enough data points to fit the model
  if(nrow(district_data) < 10) {
    return(NULL)  # Skip this district if not enough data
  }
  
  # Fit the GAM model with reduced complexity and interaction terms
  gam_model <- gam(BreedCount ~ s(KeyDateYear, bs = "cr", k = 5) + s(PrimaryBreed, bs = "re"),
                   data = district_data, method = "REML", family = poisson(link = "log"))
  
  # Predict counts for the new dataset
  new_data_district <- new_data %>% filter(DistrictSort == district)
  new_data_district$predicted_breed_count <- predict(gam_model, newdata = new_data_district, type = "response")
  
  return(new_data_district)
}

# Apply the function to each district and combine the results
predicted_data_list <- lapply(levels(new_df$DistrictSort), fit_gam_and_predict)
predicted_data <- do.call(rbind, predicted_data_list[!sapply(predicted_data_list, is.null)])

# Function to assign colors to breeds
assign_colors <- function(data) {
  breeds <- unique(data$PrimaryBreed)
  colors <- scales::hue_pal()(length(breeds))
  setNames(colors, breeds)
}
breed_color_map <- assign_colors(new_data)

# Plot the results with real data points
ggplot() +
  geom_line(data = predicted_data, aes(x = KeyDateYear, y = predicted_breed_count, color = PrimaryBreed), size = 1) +
  geom_point(data = breed_trends, aes(x = KeyDateYear, y = BreedCount, color = PrimaryBreed), size = 1) +
  facet_wrap(~DistrictSort, scales = "free_y") +
  ylim(0, 200) +
  labs(x = "Year", y = "Count", title = "Yearly Trend of Predicted and Actual Breed Counts per District") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "top") +
  guides(color = guide_legend(title = "Breed", title.position = "top"))

# Check the summary and diagnostic plots for one of the models (optional)
district_example <- levels(new_df$DistrictSort)[1]
example_model <- gam(BreedCount ~ s(KeyDateYear, bs = "cr", k = 5) + s(PrimaryBreed, bs = "re"),
                     data = breed_trends %>% filter(DistrictSort == district_example), method = "REML", family = poisson(link = "log"))
summary(example_model)
#plot.gam(example_model)


```


